---
title: "Analysis of Simulated Hungarian Income Data"
author: "Statistical Analysis Project"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: hide
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(car)
library(lmtest)
library(ggplot2)
library(dplyr)
library(broom)
library(viridis)
library(ggthemes)
library(plotly)
library(knitr)
library(kableExtra)
```

# Introduction

This analysis examines simulated income data for Hungary, focusing on the relationships between income and various demographic factors including age, location, occupation, and gender. The dataset is simulated to reflect realistic patterns while maintaining a manageable size for analysis.

```{r style_setup, include=FALSE}
# Custom theme for all plots
custom_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_line(color = "gray95")
  )

# Color palette
income_palette <- viridis::viridis(10)
```

# Data Simulation

```{r data_simulation}
set.seed(123) # For reproducibility

# Define parameters
n <- 10000  # Number of observations
cities <- c("Budapest", "Debrecen", "Szeged", "Miskolc", "Pécs", "Győr", "Szombathely", "Eger")
occupations <- c("Software Developer", "Teacher", "Doctor", "Sales Representative", 
                "Engineer", "Accountant", "Nurse", "Manager", "Chef", "Driver")
genders <- c("Male", "Female")

# Generate data
data <- data.frame(
  age = round(rnorm(n, mean = 40, sd = 12)),
  city = sample(cities, n, replace = TRUE),
  occupation = sample(occupations, n, replace = TRUE),
  gender = sample(genders, n, replace = TRUE)
)

# Ensure age is between 18 and 70
data$age <- pmax(18, pmin(70, data$age))

# Generate income based on relationships
data <- data %>%
  mutate(
    # Base income
    base_income = 300000,
    
    # Age effect (quadratic relationship)
    age_effect = 10000 * (age - 40) - 100 * (age - 40)^2,
    
    # City effect
    city_effect = case_when(
      city == "Budapest" ~ 100000,
      city %in% c("Debrecen", "Szeged") ~ 50000,
      TRUE ~ 0
    ),
    
    # Occupation effect
    occupation_effect = case_when(
      occupation == "Software Developer" ~ 150000,
      occupation == "Doctor" ~ 120000,
      occupation == "Manager" ~ 100000,
      occupation == "Engineer" ~ 80000,
      occupation == "Accountant" ~ 60000,
      occupation == "Teacher" ~ 40000,
      occupation == "Nurse" ~ 30000,
      occupation == "Chef" ~ 20000,
      occupation == "Driver" ~ 10000,
      TRUE ~ 0
    ),
    
    # Gender effect (simulating gender pay gap)
    gender_effect = ifelse(gender == "Male", 20000, 0),
    
    # Random noise
    noise = rnorm(n, 0, 50000)
  ) %>%
  mutate(
    income = base_income + age_effect + city_effect + occupation_effect + gender_effect + noise
  ) %>%
  select(age, city, occupation, gender, income)

# Save the data
write.csv(data, "hungarian_income_data.csv", row.names = FALSE)
```

# Descriptive Statistics

```{r descriptive_stats}
# Summary statistics with kable
summary_stats <- summary(data)
kable(summary_stats, caption = "Summary Statistics of the Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

# Interactive income distribution by gender
p1 <- ggplot(data, aes(x = income, fill = gender)) +
  geom_density(alpha = 0.6) +
  scale_fill_viridis_d() +
  labs(title = "Income Distribution by Gender",
       subtitle = "Density plot showing the distribution of income across genders",
       x = "Income (HUF)",
       y = "Density") +
  custom_theme

ggplotly(p1)

# Interactive income by occupation with jitter
p2 <- ggplot(data, aes(x = reorder(occupation, income, FUN = median), y = income, color = occupation)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(alpha = 0.1, width = 0.2) +
  scale_color_viridis_d() +
  coord_flip() +
  labs(title = "Income Distribution by Occupation",
       subtitle = "Boxplot with individual data points",
       x = "Occupation",
       y = "Income (HUF)") +
  custom_theme

ggplotly(p2)

# Interactive income by city with violin plot
p3 <- ggplot(data, aes(x = reorder(city, income, FUN = median), y = income, fill = city)) +
  geom_violin(alpha = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.5) +
  scale_fill_viridis_d() +
  coord_flip() +
  labs(title = "Income Distribution by City",
       subtitle = "Violin plot with boxplot overlay",
       x = "City",
       y = "Income (HUF)") +
  custom_theme

ggplotly(p3)

# Interactive age vs income scatter plot with regression line
p4 <- ggplot(data, aes(x = age, y = income, color = gender)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_color_viridis_d() +
  labs(title = "Relationship between Age and Income",
       subtitle = "Scatter plot with LOESS regression line",
       x = "Age",
       y = "Income (HUF)") +
  custom_theme

ggplotly(p4)

# Create a correlation heatmap
numeric_data <- data %>% select(age, income)
cor_matrix <- cor(numeric_data)
cor_df <- as.data.frame(cor_matrix)
cor_df$var1 <- rownames(cor_df)
cor_df_long <- pivot_longer(cor_df, -var1, names_to = "var2", values_to = "correlation")

ggplot(cor_df_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile() +
  scale_fill_viridis() +
  labs(title = "Correlation Heatmap",
       subtitle = "Correlation between numeric variables",
       x = "",
       y = "") +
  custom_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Hypothesis Testing

## Parametric Tests

### 1. Gender Income Difference (t-test)
```{r gender_ttest}
# Test if there's a significant difference in income between genders
t_test_result <- t.test(income ~ gender, data = data)
print(t_test_result)
```

### 2. City Income Differences (ANOVA)
```{r city_anova}
# Test if there are significant differences in income between cities
city_anova <- aov(income ~ city, data = data)
print(summary(city_anova))
```

### 3. Occupation Income Differences (ANOVA)
```{r occupation_anova}
# Test if there are significant differences in income between occupations
occupation_anova <- aov(income ~ occupation, data = data)
print(summary(occupation_anova))
```

## Non-parametric Tests

### 1. Gender Income Distribution (Kolmogorov-Smirnov Test)
```{r gender_ks}
# Test if income distributions differ between genders
male_income <- data$income[data$gender == "Male"]
female_income <- data$income[data$gender == "Female"]
ks_test <- ks.test(male_income, female_income)
print(ks_test)
```

### 2. Income Distribution by City (Kruskal-Wallis Test)
```{r city_kruskal}
# Test if income distributions differ between cities
kruskal_test <- kruskal.test(income ~ city, data = data)
print(kruskal_test)
```

# Regression Analysis

## Multiple Linear Regression
```{r regression}
# Convert categorical variables to factors
data$city <- as.factor(data$city)
data$occupation <- as.factor(data$occupation)
data$gender <- as.factor(data$gender)

# Fit the model
model1 <- lm(income ~ age + I(age^2) + city + occupation + gender, data = data)

# Create a beautiful model summary table
model_summary <- summary(model1)
kable(tidy(model_summary), caption = "Multiple Linear Regression Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

# Enhanced model diagnostics
par(mfrow = c(2,2))
plot(model1, col = income_palette[1], pch = 19, cex = 0.7)
```

## Polynomial Regression for Age-Income Relationship
```{r polynomial_regression}
# Fit polynomial regression
model2 <- lm(income ~ poly(age, 3), data = data)

# Create an interactive plot
p5 <- ggplot(data, aes(x = age, y = income)) +
  geom_point(alpha = 0.1, color = income_palette[1]) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), 
              color = income_palette[5], fill = income_palette[5], alpha = 0.2) +
  labs(title = "Polynomial Regression: Age vs Income",
       subtitle = "Cubic polynomial fit with confidence interval",
       x = "Age",
       y = "Income (HUF)") +
  custom_theme

ggplotly(p5)

# Model comparison
model_comparison <- data.frame(
  Model = c("Multiple Linear", "Polynomial"),
  R_squared = c(summary(model1)$r.squared, summary(model2)$r.squared),
  Adj_R_squared = c(summary(model1)$adj.r.squared, summary(model2)$adj.r.squared)
)

kable(model_comparison, caption = "Model Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

# Predictions

```{r predictions}
# Create a sample prediction
new_data <- data.frame(
  age = 35,
  city = "Budapest",
  occupation = "Software Developer",
  gender = "Male"
)

# Predict income
prediction <- predict(model1, newdata = new_data, interval = "prediction")
print(prediction)
```

# Summary and Conclusions

The analysis of the simulated Hungarian income data reveals several interesting patterns:

1. There is a significant gender pay gap, with males earning more on average than females.
2. Income varies significantly across different cities, with Budapest showing the highest average income.
3. Occupation has a strong effect on income, with software developers and doctors earning the most.
4. Age shows a non-linear relationship with income, peaking in the middle age range.
5. The regression models explain a significant portion of the income variation.

The analysis demonstrates the complex interplay between various factors affecting income levels in Hungary. The findings suggest that both demographic factors (age, gender) and professional factors (occupation, location) play important roles in determining income levels.

# References

- R Core Team (2023). R: A language and environment for statistical computing.
- Wickham, H. (2016). ggplot2: Elegant Graphics for Data Analysis.
- Fox, J. (2016). Applied Regression Analysis and Generalized Linear Models. 